---
title: "profiling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{profiling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Synopsis
* TCA is a great package. But it is running unexpectedly slowly. A project with
a moderately sized (400,000 CpGs * 1000 Samples * 6 Cell Types, vars.mle = FALSE)
couldn't converge after 12 hours and hit out of memory error with 120GB of RAM.
* This fork is created to profile the performance of TCA and see if there are 
bottlenecks we can resolve.  

```{r setup, eval = FALSE}
library(profvis) # Profiling tool
library(devtools)
library(tictoc)
load_all()

# Simulate the data
set.seed(1234)
data <- test_data(400000, 20, 6, 1, 1, 0.01)
lapply(data, dim)

set.seed(1234)
# Run the data sequentially first.
sequential <- profvis({
  tca.mdl <- tca(X = data$X, W = data$W, C1 = data$C1, C2 = data$C2)
})

sequential
```

* The sequential run took 54s to be done
* We see that in `model_fit.R/tca_fit.R`, r is spending 
  + High peak memory (1.4GB) and decent amount of time (6s) to call `tca.fit_means_vars`. 
  This is the internal loop convergence of TCA. This is probablly what 
  bottlenecked the loop in parallel mode because the high memory overhead.  
  + High peak memory (20 GB) and a lot of time (30s) to call `data.frame()`.
  before `lm()` to calculate the `gammas_hat_pvals.joint`.  
* Let's try to replace the `data.frame()` call with a straight `cbind()` call to 
create a X matrix. Then we can use `RcppEigen::fastLm()` to directly call 
the data matrix and the vector y.  
* This is potentially a dangerous trade-off between speed and safety. But 
since NA is already expected to be taken care of by user, we should be safe to 
make this trade off. Other issues such as full rank matrix should still cause 
errors. We can create a test case for this.  

# Check lm functions
```{r, eval = FALSE}
lm_model <- lm(cyl ~ mpg + hp + disp, data = mtcars)
fastlm_model <- RcppEigen::fastLm(
  cbind(1, as.matrix(mtcars[, which(names(mtcars) %in% c("mpg", "hp", "disp"))])),
  mtcars[, "cyl"]
)
```

* Results are almost identical
```{r}
summary(lm_model) |> names()
summary(lm_model)$coefficients
summary(fastlm_model)$coefficients
```

```{r, eval = FALSE}
# para <- profvis({
#   tca.mdl <- tca(X = data$X, W = data$W, C1 = data$C1, C2 = data$C2,
#                  parallel = TRUE, num_cores = 3)
# })

df <- data.frame(
  y = X_tilde[, j],
  cbind(
    W / t(repmat(W_norms[, j], k, 1)),
    if (p2 > 0) {
      C2 / t(repmat(W_norms[, j], p2, 1))
    } else {
      C2
    },
    C1_null
  )
)

tca_fit_debug <- qs::qread("./data-raw/tca_fit_profile.qs")
tca_fit_debug$t1

profvis({
  test <- cbind(tca_fit_debug$t1, tca_fit_debug$t2)
  df <- data.frame(test)
})

test <- lm(vs ~ mpg, mtcars)
test$model

model.matrix(vs ~ mpg, data = mtcars)
mtcars[, "mpg", drop = FALSE]
mtcars[, "vs"]

cbind(1, mtcars[, "mpg", drop = FALSE])

lm.fit(cbind(1, as.matrix(mtcars[, "mpg", drop = FALSE])), mtcars[, "vs"])
Rfast::lmfit()
RcppEigen::fastLm(
  X = cbind(1, as.matrix(mtcars[, "mpg", drop = FALSE])),
  y = mtcars[, "vs"]
) |> summary()

# insert into model_fit.R/tca.fit.R
message("saving profile")
qs::qsave(
  list(t1 = t1, t2 = t2, p2 = p2, C2 = C2, W_norms = W_norms),
  "./data-raw/tca_fit_profile.qs"
)
```

